Synthesis of Linear Functional Programs

### 2021-07-20 article/article.tex

\mypara{Refinement Types}

@@

\mypara{General} The implementation was carried out in Haskell. We
highlight some key points.

\mypara{Backtracking} Even though focusing reduces non-determinism
during proof search, it cannot fully eliminate it. At several points
during proof construction we must choose one of multiple rules to
apply, and in the event of not being able to construct the proof
following that choice, we want to backtrack to the decision point and
attempt to derive a proof taking a different route. To this effect,
our prototype synthetiser depends on \emph{LogicT}, "a backtracking
logic-programming monad"~\cite{logict}.

@@

\mypara{Memoization} Some proofs attempt to synthetise the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthetise result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values and
uses the result as the key\todo{via a hashconsing library?}.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take
around $5s$, while synthesis without memoization doesn't terminate
even after $15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.

>>>>

\mypara{Refinement Types} Their syntax is... Their rules are...

\begin{mathpar}
    \infer*[right=(refR)]
    {  }
    {  }
\end{mathpar}

@@

The implementation was carried out in Haskell. We
highlight some key points.

\mypara{Backtracking} Even though focusing reduces non-determinism
during proof search, it cannot fully eliminate it. At several points
during proof construction we must choose one of multiple rules to
apply, and in the event of not being able to construct the proof
following that choice, we want to backtrack to the decision point and
attempt to derive a proof taking a different route. To this effect,
our prototype synthetiser depends on \emph{LogicT}, "a backtracking
logic-programming monad"~\cite{logict}. Special attention is
required for ADTs. When deconstructing an ADT, since all branches must be
synthetisable and result in the same ouput linear context $\Delta'$, alternative
constructor deconstructions should be considered fairly in conjunction, so the
\textbf{fair conjunction} operator should be used. When constructing an ADT, the
disjunctive choice of constructor should be fair, and so the \textbf{fair
disjunction} operator used. Fairness at these two points usually increases
the synthesis speed slightly, but can be a relevant factor in synthesis in which a
combination of choices following the ADT deconstruction or construction can
backtrack many times or even arbitrarily.

@@

\mypara{Memoization} Some proofs attempt to synthetise the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthetise result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values and
uses the result as the key\todo{via a hashconsing library?}.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take
around $5s$, while synthesis without memoization doesn't terminate
even after $15m$ running.
>>>>>>> 4290cfe31b8de10ca70f772fbc7d5fdc9542273c
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.
